Light and life form interact to shape stomatal ratio among British angiosperms

### 2017-07-14 ms/ms.tex
reran everything to check
edited methods

\begin{itemize}
	\item In most plants, stomata are located only on the abaxial leaf surface (hypostomy), but many plants have stomata on both surfaces (amphistomy). High light and herbaceous growth form have been hypothesized to favor amphistomy, but these hypotheses have not been rigourously tested together using phylogenetic comparative methods.
	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, Raunki\ae r lifeform, and phylogenetic relationships for 372 species of British angiosperms. I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio.
	\item High light and herbaceous growth form are correlated with amphistomy, as predicted, but they also interact; the effects of light are most pronounced in therophytes (annuals) and hemocryptophytic perennials, but less so in phanerophytes (mostly trees). Interestingly, amphistomy and stomatal density evolve together in response to light, suggesting coordinated selection on this trait combination.
	\item I show for the first time that light and growth form interact to shape variation in stomatal ratio; amphistomy is advantageous in high light, but mostly for herbs. These results improve our understanding of the adaptive significance of stomatal ratio as well as its use as functional trait for paleovegetation reconstruction and crop improvement.
\end{itemize}

@@

% NOTES FOR FIRST TWO PARAGRAPHS
% Stomatal ratio and photo/water use: Parkhurst papers, 
% from Peat and Fitter 1994b "Jones (1985) found that amphistomaty may be advantageous as it allows greater CO2 transfer through the mesophyll, especially in thicker leaves, and greater leaf conductance in high light conditions when CO2 is limiting photosynthetic rate. However, his models showed that hypostomaty would be advantageous when there are high boundary layer resistances (for example in large leaves or deep within a plant canopy), high humidity or a temperature gradient across the leaf."

%%%%% PARAGRAPH 3-4: adaptive hypotheses and current limitations %%%%%
The leading, nonmutually exclusive hypotheses are that 1) open habitats favour amphistomy because CO$_2$ diffusion most strongly limits photosynthetic rate under high light and 2) herbaceous growth form favours amphistomy because traits that maximize photosynthetic rate are often under stronger selection in herbs. \citeauthor{Salisbury_1927} (\citeyear{Salisbury_1927}) first noted that amphistomy is most common in herbaceous plants from open habitats (i.e., with high light) of the British flora. These observations have been replicated in other studies \citep{Mott_etal_1982, Peat_Fitter_1994b, Jordan_etal_2014, Muir_2015} and may support physiological and ecological hypotheses that CO$_2$ most strongly limits photosynthesis in high light and/or photosynthesis contributes most to fitness in herbaceous plants. Under high light, CO$_2$ can strongly limit maximum photosynthetic rates, espcecially in thick leaves \citep{Jones_1985}. Hence, having stomata on both surfaces relieves this limitation by adding a second parallel pathway for CO$_2$ diffusion. Parkhurst \citeyear{Parkhurst_1978} argued that greater leaf thickness \textit{per se} selected for amphistomy, but there is little evidence for correlations between leaf thickness and stomatal ratio independent of light \citep{Mott_etal_1982, Gibson_1996, Muir_2015}. Amphistomy is correlated with open habitat in warm desert plants of western North America \citep{Mott_etal_1982, Gibson_1996}, among the Proteaceae \citep{Jordan_etal_2014}, and in continental European herbs \citep{Bucher_etal_2017}.

@@

% NOTES FOR PARAGRAPHS 3-4:
% From Mott et al 1982: ''Korner, Scheel \& Bauer (1979) have recently reviewed the literature on maximum stomatal conductance and photosynthetic rates for plants of specific habitats and growth forms. Low maximum photosynthetic rates are observed for succulents, evergreen conifers, deciduous woody plants, herbs from shady habitats, evergreen woody plants, and desert and steppe shrubs. Higher maximum values are obtained for deciduous fruit trees, wild grasses, cultivated C$_4$ and C$_3$ grasses, herbaceous crop plants, herbs from open habitats, and plants from aquatic habitats and swamps. Bazzaz (1979) cites evidence that the maximum photosynthetic rate declines through successional stages in a plant community, pointing to higher rates for open field plants and grasses and lower rates for canopy deciduous trees and shade habitat herbs. Early successional trees appear to be intermediate in photosynthetic rates.''

%%%% PARAGRAPH 5-6 %%%%%
Although previous comparative studies have tested whether open habitat and growth form influence stomatal ratio, we do not know if these effects are independent of one another. Open habitat and growth form may not be independent because open habitats generally consist of more short-statured, herbaceous plants. Some authors have attempted to disentangle light and growth form by contrasting herbs from open and understory habiats \citep{Salisbury_1927}. However, this is problematic if phylogenetic relationships are not controlled for, because shade species may share traits simply because they are more closely related to each other than they are to high light species. Finally, open habitat and groth form may also interact with one another. For example, amphistomy may only be favored when CO$_2$ strongly limits photosynthetic rate (e.g. in high light) \textit{and} photosynthetic rate strongly limits fitness (e.g. in herbs).

@@

% from previous paper
% Here, we examine whether stomata on the abaxial (`lower') surface of the leaf evolve independently of adaxial (`upper') stomata. Stomata are microscopic pores on the leaf surface formed by a pair of guard cells. The density, size, and arrangement of stomata on a leaf set the maximum stomatal conductance to CO$_2$ diffusing into a leaf and the amount of water that transpires from it \citep{Parkhurst_1978, Sack_etal_2003, Franks_Farquhar_2001, Galmes_etal_2013}. Hence, stomatal traits like density, size, and ratio of upper to lower stomata have strong effects on carbon assimilation and water-use efficiency.

% The proportion of stomata found on the upper surface also tends to increase during domestication, even as the total stomatal density stays constant \citep{Milla_etal_2013}. Amphistomy increases CO$_2$ diffusion within the leaf by opening up a second parallel pathway in the intercellular airspace for diffusion from substomatal cavities to mesophyll cell walls. However, stomata on the upper surface in particular may be costly. For example, upper stomata increase the susceptibility to rust pathogens in \textit{Populus} \citep{Mckown_etal_2014}. Amphistomy may also cause the palisade mesophyll to dry out under strong vapor pressure deficits \citep{Buckley_etal_2015}. \cite{Muir_2015} reviewed the literature on other possible fitness costs.

@@

I obtained data on ab- and adaxial stomatal density on 395 species from British Ecological Flora \citep{Salisbury_1927, Fitter_Peat_1994a, BEF}. Following recent comparative analyses \citep[e.g.][]{Bartelheimer_Poschlod_2016, Salguero-Gomez_etal_2016}, I used Ellenberg light indicator values \citep{Ellenberg_1974} and Raunki\ae r life form \citep{Raunkiaer_1934} as measures of light habitat and growth form, respectively. Hence, I am assuming that the species' light habitat is closely related to the type of habitat (open versus closed) where that species is found. Both attributes have been recently updated by taxonomic experts of the British flora (PLANTATT, \cite{Hill_etal_2004}). Ellenberg light indicator values are hereafter abbreviated \el. I used a dated molecular phylogeny of the British flora \citep{Lim_etal_2014} available from TreeBASE (http://treebase.org/; accession number 15105). 14 species (3.5\%) in the dataset were not present in the phylogeny. For 8 of these species, I used the position a congeneric species as a proxy for the focal species. When multiple congeneric species were present, I consulted the phylogenetic literature to identify the most closely related proxy species \citep{Scheen_etal_2004, Salmaki_etal_2013}. For the remaining 6 missing species, I positioned them in the tree based on phylogenetic relationships to other genera or families present in the tree \citep{Fior_etal_2006}. Because many phylogenetic comparative methods do not allow polytomies, zero-length branches, and non-ultrametric trees, I made several small adjustments to the tree. I resolved polytomies randomly using the `multi2di' function in \pkg{phytools} version 0.5-64 \citep{Revell_2012}. I added 0.02 my to all zero-length branches, as this was approximately the length of the shortest nonzero branch length in the tree. After these changes, I slightly altered terminal branch lengths to make the tree precisely ultrametric.

>>>>

\begin{itemize}
	\item In most plants, stomata are located only on the abaxial leaf surface (hypostomy), but many plants have stomata on both surfaces (amphistomy). High light and herbaceous growth form have been hypothesized to favor amphistomy, but these hypotheses have not been rigourously tested together using phylogenetic comparative methods.
	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, Raunki\ae r lifeform, and phylogenetic relationships for 372 species of British angiosperms. I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio.
	\item High light and herbaceous growth form are correlated with amphistomy, as predicted, but they also interact; the effects of light are most pronounced in therophytes (annuals) and perennial herbs, but less so in phanerophytes (mostly trees). Interestingly, amphistomy and stomatal density evolve together in response to light, suggesting coordinated selection on this trait combination.
	\item I show for the first time that light and growth form interact to shape variation in stomatal ratio; amphistomy is advantageous in high light, but mostly for herbs. These results improve our understanding of the adaptive significance of stomatal ratio as well as its use as functional trait for paleovegetation reconstruction and crop improvement.
\end{itemize}

@@

The leading, nonmutually exclusive hypotheses are that 1) open habitats favour amphistomy because CO$_2$ diffusion most strongly limits photosynthetic rate under high light and 2) herbaceous growth form favours amphistomy because traits that maximize photosynthetic rate are often under stronger selection in herbs. \citeauthor{Salisbury_1927} (\citeyear{Salisbury_1927}) first noted that amphistomy is most common in herbaceous plants from open habitats (i.e., with high light) of the British flora. These observations have been replicated in other studies \citep{Mott_etal_1982, Peat_Fitter_1994b, Jordan_etal_2014, Muir_2015} and may support physiological and ecological hypotheses that CO$_2$ most strongly limits photosynthesis in high light and/or photosynthesis contributes most to fitness in herbaceous plants. Under high light, CO$_2$ can strongly limit maximum photosynthetic rates, espcecially in thick leaves \citep{Jones_1985}. Hence, having stomata on both surfaces relieves this limitation by adding a second parallel pathway for CO$_2$ diffusion. Parkhurst \citeyear{Parkhurst_1978} argued that greater leaf thickness \textit{per se} selected for amphistomy, but there is little evidence for correlations between leaf thickness and stomatal ratio independent of light \citep{Mott_etal_1982, Gibson_1996, Muir_2015}. Amphistomy is correlated with open habitat in warm desert plants of western North America \citep{Mott_etal_1982, Gibson_1996}, among the Proteaceae \citep{Jordan_etal_2014}, and in continental European herbs \citep{Bucher_etal_2017}.

@@

Although previous comparative studies have tested whether open habitat and growth form influence stomatal ratio, we do not know if these effects are independent of one another. Open habitat and growth form may not be independent because open habitats generally consist of more short-statured, herbaceous plants. Some authors have attempted to disentangle light and growth form by contrasting herbs from open and understory habiats \citep{Salisbury_1927}. However, this is problematic if phylogenetic relationships are not controlled for, because shade species may share traits simply because they are more closely related to each other than they are to high light species. Finally, open habitat and groth form may also interact with one another. For example, amphistomy may only be favored when CO$_2$ strongly limits photosynthetic rate (e.g. in high light) \textit{and} photosynthetic rate strongly limits fitness (e.g. in herbs).

@@

@@

I obtained data on ab- and adaxial stomatal density on 395 species from British Ecological Flora \citep{Salisbury_1927, Fitter_Peat_1994a, BEF}. Following recent comparative analyses \citep[e.g.][]{Bartelheimer_Poschlod_2016, Salguero-Gomez_etal_2016}, I used Ellenberg light indicator values \citep{Ellenberg_1974} and Raunki\ae r life form \citep{Raunkiaer_1934} as measures of light habitat and growth form, respectively. Hence, I am assuming that the species' light habitat is closely related to the type of habitat (open versus closed) where that species is found. Both attributes have been recently updated by taxonomic experts of the British flora (PLANTATT, \cite{Hill_etal_2004}). Ellenberg light indicator values are hereafter abbreviated \el. I used a dated molecular phylogeny of the British flora \citep{Lim_etal_2014} available from TreeBASE (http://treebase.org/; accession number 15105). 14 species (3.5\%) in the dataset were not present in the phylogeny. For 8 of these species, I used the position a congeneric species as a proxy for the focal species. When multiple congeneric species were present, I consulted the phylogenetic literature to identify the most closely related proxy species \citep{Scheen_etal_2004, Salmaki_etal_2013}. For the remaining 6 missing species, I positioned them in the tree based on phylogenetic relationships to other genera or families present in the tree \citep{Fior_etal_2006}. Because many phylogenetic comparative methods do not allow polytomies, zero-length branches, and non-ultrametric trees, I made several small adjustments to the tree. I resolved polytomies randomly using the `multi2di' function in \pkg{phytools} version 0.6-00 \citep{Revell_2012}. I added 0.02 my to all zero-length branches, as this was approximately the length of the shortest nonzero branch length in the tree. After these changes, I slightly altered terminal branch lengths to make the tree precisely ultrametric.
